### 2013-03-28 doc/sosp13/evaluation.tex
new eval. commit to merge

\section{Evaluation} \label{sec:evaluation}
Our evaluation begins with presentation of microbenchmark results, establishing
the intrinsic potential of Grappa to provide random access bandwidth and latency tolerance.
Next, we present application results, both for Grappa and for other paradigms, as well
as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
on Grappa results, thus exploring robustness to network scale.

\subsection{Microbenchmark Results}

\subsubsection{Random Access}
\TODO{Random access feed forward results on Grappa.  optional: Results we measured for XMT.  Cite MPI results}
\subsubsection{Latency Tolerance}
\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on Grappa.}
\subsubsection{Scheduling and Robustness}

@@

\TODO{Summary of what to write please expand: Reference above plots. Bandwidth of single socket is
    270Mcacheline/s. Each context is 4 cachelines: 1 for worker struct
    and 3 stack cachelines. We must read and write every context, so 8
    cacheline transfers per context switch. This asymptotically
    approaches 33.75Mcontexts/s as you increase the number of contexts
    (fewer and fewer in L3). Note that only 4+ cores reaches full rate
    because these westmere chips are balanced to not achieve full off-chip bandwidth until 4
    cores--bdmyers}

\TODO{Yield test results:  latency \& bandwidth.  Discussion of implications of zillions of contexts on robustness of latency tolerance preshadowing ~\ref{sec:scaling}}

>>>>

\section{Evaluation} \label{sec:evaluation}

The goal of our evaluation is to show that the core pieces of the \Grappa
runtime system, namely our tasking system and the global memory/communication
layer work as expected and together are able to efficiently run irregular
applications. We evaluate \Grappa in three basic steps:
\begin{itemize}
\item We present results
that show that the \Grappa runtime is able to sustain very high concurrency
rates and the communication layer is able to sustain a very high rate of
global memory operations. We also show the performance of a graph kernel that
stresses communication and concurrency together.

\item We show how some popular irregular applications running on \Grappa compare to the Cray XMT and hand-tuned MPI code.

\item We finish with a characterization of system behavior, including profiling where execution time goes, how aggregation affects message size and rates, how global memory and work stealing behaves.

\end{itemize}

\subsection{Basic Grappa mechanisms}

\subsection{Fast User-Level Context Switching }

@@

% More specifically, each context is cache lines: 1 for the worker struct and 3
% for stack data. So the are so 8 cacheline transfers per context switch (write
% the old context, read the next in). The bandwidth of single socket is 270M cache lines/s.
% 
% This asymptotically approaches
% 33.75Mcontexts/s as you increase the number of contexts (fewer and fewer in
% L3). Note that only 4+ cores reaches full rate because these westmere chips
% are balanced to not achieve full off-chip bandwidth until 4 cores.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% OLD %%%%%%%%%%%%%%%%%%

Our evaluation begins with presentation of microbenchmark results, establishing
the intrinsic potential of Grappa to provide random access bandwidth and latency tolerance. 
Next, we present application results, both for Grappa and for other paradigms, as well
as comparing against the Cray XMT.  Finally, we present the impact of increased aggregation delay
on Grappa results, thus exploring robustness to network scale.

\subsection{Microbenchmark Results}
\subsubsection{Random Access}
\TODO{Random access feed forward results on Grappa.  optional: Results we measured for XMT.  Cite MPI results}
\subsubsection{Latency Tolerance}
\TODO{Simple ping test results -- eg, MPI ping, not the full blown aggregation ping test.  Random access blocking results on Grappa.}
\subsubsection{Scheduling and Robustness}
\TODO{Yield test results:  latency \& bandwidth.  Discussion of implications of zillions of contexts on robustness of latency tolerance preshadowing ~\ref{sec:scaling}}

### 2013-03-28 doc/sosp13/intro.tex
new eval. commit to merge

\noindent{\bf Poor data locality and frequent communication.} Data reference
patterns of irregular applications are unpredictable and tend to be spread
across the entire system. This results in frequent requests for remote data.
Caches are of little assistance because of low data re-use and spatial
locality. Prefetching is ineffective because request locations are not known
early enough. Data-parallel frameworks such as
MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on
being able to partition data and regular communication patterns. As a
consequence, irregular applications require frequent communication of
typically small pieces of data. In contrast, commodity networks are designed
for large packets and have just a fraction of their peak bandwidth when moving
small packets.

\noindent{\bf High network communication latency.} The performance challenges
of frequent communication is exacerbated by the network latency relative to
processor performance. Latency of commodity networks runs anywhere from one to
hundreds of microseconds, which is much more than superscalar execution with
sophisticated memory hierarchies can tolerate. Since irregular application
tasks encounter remote references dynamically during execution and must
resolve them before making further progress, stalls are frequent and lead
severely underutilized compute resources.

While some irregular applications can be restructured to better exploit
locality, aggregate requests to increase network message size, and manage the
additional challenges of load balance and synchronization across multinode
systems, the effort required to do so is formidable and involves knowledge and
skills pertaining to distributed systems far beyond those of most application
programmers. Luckily, many of the important irregular applications (e.g.,
graph processing, our focus in this paper) naturally offer large amounts of
concurrency. This immediately suggests taking advantage of concurrency to
tolerate the latency of data movement by overlapping computation with
communication.

>>>>

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication.} Data reference
patterns of irregular applications are unpredictable and tend to be spread
across the entire system. This results in frequent requests for remote data.
Caches are of little assistance because of low data re-use and spatial
locality. Prefetching is ineffective because request locations are not known
early enough. Data-parallel frameworks such as
MapReduce~\cite{mapreduce:osdi04} are also inappropriate because they rely on
being able to partition data and regular communication patterns. As a
consequence, irregular applications require frequent communication of
typically small pieces of data. In contrast, commodity networks are designed
for large packets and have just a fraction of their peak bandwidth when moving
small packets.

\vspace{0.5ex}
\noindent{\bf High network communication latency.} The performance challenges
of frequent communication is exacerbated by the network latency relative to
processor performance. Latency of commodity networks runs anywhere from one to
hundreds of microseconds, which is much more than superscalar execution with
sophisticated memory hierarchies can tolerate. Since irregular application
tasks encounter remote references dynamically during execution and must
resolve them before making further progress, stalls are frequent and lead
severely underutilized compute resources.

While some irregular applications can be restructured to better exploit
locality, aggregate requests to increase network message size, and manage the
additional challenges of load balance and synchronization, the effort required
to do so is formidable and involves knowledge and skills pertaining to
distributed systems far beyond those of most application programmers. Luckily,
many of the important irregular applications (e.g., graph processing, our
focus in this paper) naturally offer large amounts of concurrency. This
immediately suggests taking advantage of concurrency to tolerate the latency
of data movement by overlapping computation with communication.

### 2013-03-28 doc/sosp13/relatedwork.tex
new eval. commit to merge

% Grappa is built on many existing ideas in programming languages, systems and
% architecture. In this section we discuss related frameworks and key enabling
% technologies that Grappa builds upon.

\paragraph{Distributed graph processing} While \Grappa is a general runtime
system for any large-scale concurrent application, one of its killer
applications is graph analysis. Therefore, distributed graph processing
frameworks like Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12} are related. Pregel adopts a
bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
workloads that could prioritize vertices. GraphLab overcomes this limitation
with an execution mode that schedules vertex computations individually,
allowing prioritization, which gives faster convergence in a variety of
iterative algorithms. GraphLab, however, imposes a rigid computation model
where programmers must express computation as transformations on a vertex and
its edge list only, with information only from adjacent vertexes. Pregel is
only slightly less restrictive, as the input data can be any vertex in the
graph. \Grappa also supports dynamic parallelism with asynchronous execution,
but parallelism is expressed as tasks or loop iterations, which is a far more
general programming model for irregular computation tasks.
PowerGraph~\cite{powergraph:osdi12} improves the performance of GraphLab for
real-world graphs with power-law degree distributions by using a vertex cut
for graph partitioning. If the size of the partial result of applying a subset
of edge data to the vertex is constant, then there is a good bound on the
communication required to keep vertex replicas coherent. We do not explore
these algorithmic techniques in this work, but they are complimentary and
would improve the performance of graph applications on \Grappa.

\paragraph{Software distributed shared memory}
% Grappa supports a global memory namespace, presenting a programming
% abstraction similar to that of previous software DSM systems but with
% differences attuned to irregular application demands.
Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
and aim to hide the fact that they are built in software from applications by
exploiting the processor's paging mechanisms, therefore relying heavily on
locality. Instead, \Grappa, like partitioned global address space (PGAS)
models, implements its DSM at the language, rather than the system level.
Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
UPC~\cite{upc:2005} make accesses to shared structures look like normal
references. As we describe later, \Grappa chooses a middle ground, where global
addresses are explicit in the API and local accesses are emitted
conventionally by the compiler.

Past DSM work was mostly page-based, so it could directly take advantage of a
network's RDMA support to move large page-sized blocks of data from node to
node \TODO{citation}. However, when commodity networks move small blocks of
data (a few bytes), only a fraction of the available bandwidth is achieved. \Grappa had to include significant aggregation support to make effective use of RDMA.

In summary, while \Grappa's DSM system is conceptually similar to prior work,
we accept the random access aspect of irregular applications and optimize for
throughput rather than low latency. Our DSM system must support enough memory
concurrency to tolerate the latency of the network.

% Importantly, global memory support in Grappa
% ends up being tightly coupled to the task scheduler in order to overlap long
% latency memory operations with useful computation. For these two reasons it
% became necessary to build a new DSM system specifically for Grappa.

%% We could be stronger here on why Grappa's task library is different,
%% but we are running out of time.  -Mark

\paragraph{Multithreading} \Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, \Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required to tolerate
the very high inter-node communication latency of commodity networks. \Grappa's
task library employs several optimizations: an extremely fast task switch, a
small task size, and judicious use of software prefetch instructions to bring task state
into the cache sufficiently long before that task is actually scheduled. The main difference
between \Grappa's support for lightweight threads and prior work such as
QThreads~\cite{Wheeler08qthreads:an} and
Capriccio~\cite{Behren03capriccio:scalable} is the use of prefetching by the scheduler, which is needed for good performance
when multiplexing such a large number of tasks.

>>>>

\vspace{0.5ex}
\noindent{\bf Distributed graph processing} While \Grappa is a general runtime
system for any large-scale concurrent application, one of its killer
applications is graph analysis. Therefore, distributed graph processing
frameworks like Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12} are related. Pregel adopts a
bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
workloads that could prioritize vertices. GraphLab overcomes this limitation
with an execution mode that schedules vertex computations individually,
allowing prioritization, which gives faster convergence in a variety of
iterative algorithms. GraphLab, however, imposes a rigid computation model
where programmers must express computation as transformations on a vertex and
its edge list only, with information only from adjacent vertexes. Pregel is
only slightly less restrictive, as the input data can be any vertex in the
graph. \Grappa also supports dynamic parallelism with asynchronous execution,
but parallelism is expressed as tasks or loop iterations, which is a far more
general programming model for irregular computation tasks.
PowerGraph~\cite{powergraph:osdi12} improves the performance of GraphLab for
real-world graphs with power-law degree distributions by using a vertex cut
for graph partitioning. If the size of the partial result of applying a subset
of edge data to the vertex is constant, then there is a good bound on the
communication required to keep vertex replicas coherent. We do not explore
these algorithmic techniques in this work, but they are complimentary and
would improve the performance of graph applications on \Grappa.

\vspace{0.5ex}
\noindent{\bf Software distributed shared memory.} Many traditional software
DSM systems are page based~\cite{Treadmarks,munin} and aim to hide the fact
that they are built in software from applications by exploiting the
processor's paging mechanisms, therefore relying heavily on locality. Instead,
\Grappa, like partitioned global address space (PGAS) models, implements its
DSM at the language, rather than the system level. Languages such as
Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and UPC~\cite{upc:2005}
make accesses to shared structures look like normal references. As we describe
later, \Grappa chooses a middle ground, where global addresses are explicit in
the API and local accesses are emitted conventionally by the compiler. Past
DSM work was mostly page-based, so it could directly take advantage of moving
large page-sized blocks of data from node to node. However, when commodity
networks move small blocks of data (a few bytes), only a fraction of the
available bandwidth is achieved. \Grappa had to include significant
aggregation support to make effective use of of the network.

In summary, while \Grappa's DSM system is conceptually similar to prior work,
we accept the random access aspect of irregular applications and optimize for
throughput rather than low latency. Our DSM system must support enough memory
concurrency to tolerate the latency of the network and additional latency
overhead impose by the runtime system.

\vspace{0.5ex}
\noindent{\bf Multithreading} \Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, \Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required to tolerate
the very high inter-node communication latency of commodity networks.
\Grappa's task library employs several optimizations: an extremely fast task
switch, a small task size, and judicious use of software prefetch instructions
to bring task state into the cache sufficiently long before that task is
actually scheduled. The main difference between \Grappa's support for
lightweight threads and prior work such as
QThreads~\cite{Wheeler08qthreads:an} and
Capriccio~\cite{Behren03capriccio:scalable} is context prefetching, which is needed for good performance when multiplexing such a large
number of tasks.

### 2013-03-28 doc/sosp13/tasks.tex
moved graphs
new eval. commit to merge

The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
a single operating system thread pinned to it. When tasks are ready to
execute, they are mapped to a {\em worker}, which is a user-level thread.
\Grappa keeps a large number of workers per hardware core, which is at the
core of \Grappa's latency tolerance properties. Below we discuss the
implementation of our task management support and then describe how
applications expose parallelism to the \Grappa runtime.

@@

\paragraph{Workers}

When a new task is spawned, it is enqueued until it is ready for execution.
Tasks are not allocated any execution resources (a worker) until the scheduler
decides to run them. A worker is simply a collection of status bits and a
stack, allocated at a particular core. There are two types of workers: (1) user worker; and (2) system worker.

@@

\paragraph{Scheduling}

During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the \Grappa scheduler operates entirely in user-space
and does little more than store register state of one task and load that of
another.

@@

Each core in a \Grappa system has its own independent scheduler. Each
scheduler has three main responsibilities: servicing system workers that
process communication requests as described in
Section~\ref{sec:communication}; rescheduling workers that have been waiting
on long-latency operations; and assigning ready tasks to worker resources that
have become idle.

@@

\paragraph{Context switching} \Grappa context switches between workers
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{amd64:abi:2012}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save, according to the ABI, is
62~bytes.

%% No attempt is made to fit them within a single cache line unfortunately :-( Mark
%%% --- importantly, a context is saved in a single 64 byte cache line.

@@

Since \Grappa keeps a very large number of active workers, their context data will not fit in cache. Therefore, the scheduler has to perform very careful prefetching in order to reduce context switch cost to a minimum. We determined empirically that prefetching the fourth worker in the scheduling order is sufficient. Also determined empirically, we prefetch three cache lines from the stack for each worker.  This makes context switching effectively free of cache misses even to hundreds of thousands of threads.

% When the scheduler chooses a thread to schedule it first looks to see if it must run a periodic task.  If so it choose that.  If not it looks for a ready task.  If none exists it chooses the idle thread.
% 
% If it chooses a task from the ready queue it prefetches the 4'th next thread on the read queue.  Presently 192 bytes of stack are prefetched and 64 bytes of the thread struct itself.
% 
% A periodic task is responsible for looking at the size of the ready queue.  If it is too small more not started tasks are started.
%

% The context switcher is written entirely in user space.  A context is saved in a single cache line (64 bytes). We prefetch 192 bytes from the stack (which is where the context is saved) in order to prime the stride prefetched on the core and to preload the task state that exists on the stack.

% Fast context switching amongst many tasks is crucial to Grappa for two
% reasons. First, only when the net cost of context switching is dominated by
% the loss incurred by stalling does a latency tolerant approach work to
% increase throughput on computations with intense random access. Random access
% latencies to node local memory are on the order of 100ns. Were the scheduler
% to add more overhead than this, it would actually slow down rather than speed
% up computations that randomly access node local data. Second, because we can
% switch amongst so many tasks without dramatically increasing scheduling
% overhead, Grappa can tolerate system latencies many orders of magnitude
% greater than that of local memory. Network ping latency alone is approximately
% 2us, but our data suggests we can tolerate latencies in the millisecond scale.
% Furthermore, by over-provisioning task parallelism, Grappa promises to provide
% robust throughput that is relatively insensitive to variations in latencies
% within and across our target systems.

@@

\comment{Commenting out.  Simon things it doesn't add much I worried about its usefulness too.  -M

\TODO{Anything after this I'm not sure adds to the paper -- if it does, please edit -Simon}
The
core would be fully-utilized if there is always something on the ready
queue. Even if there are many active tasks, if they are all suspended
for long-latency network requests, then the core is underutilized. So,
there is a choice: should the core use some of this idle time to perform
work stealing or should it just wait as it is possible that local tasks
will create more work. We choose to have a core initiate a steal when
all of the following conditions hold: no workers are ready to run, the
unstarted task queues are empty, and there are no outstanding steal
requests. Having only one outstanding steal request throttles steals to
prevent flooding the network, but other scalable quieting mechanisms
would be possible, such as a voting
tree\cite{scalableWorkStealingOrCilk98} or lifelines \cite{lifelines}.
Termination detection is not built into the \Grappa task scheduler,
rather, it is considered a programming error for the program to exit
without syncing all tasks.
}

@@

\begin{figure}[htbp]
  \begin{center}
    \begin{description}\small
    \item[ \texttt{spawn( void (*fp)(args) )} ] \hfill \\
      Creates a new stealable task
    \item[ \texttt{spawn\_on( core, (*fp)(args) )} ] \hfill \\
      Creates a new private task that will run on a specific core 
    \item[ \texttt{parallel\_for( (*fp)(args), start, end )} ] \hfill \\
      Executes iterations of a loop as stealable tasks 
    \item[ \texttt{call\_on( core, (*fp)(args) )} ] \hfill \\ 
      Runs a limited function on a specific core without consuming
      \Grappa execution resources 
    \end{description}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
      } % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}

>>>>

The basic unit of execution in \Grappa is a {\em task}. Each hardware core has
a single operating system thread pinned to it. When tasks are ready to
execute, they are mapped to a {\em worker}, which is akin to a user-level
thread. \Grappa keeps a large number of workers per hardware core, this is the
core reason why \Grappa is able to tolerate latencies well. Below we discuss
the implementation of our task management support and then describe how
applications expose parallelism to the \Grappa runtime.

@@

\paragraph{Workers}
When a new task is spawned, it is enqueued until it is ready for execution.
Tasks are not allocated any execution resources (a worker) until the scheduler
decides to run them. A worker is simply a collection of status bits and a
stack, allocated at a particular core. There are two types of workers: (1) user worker; and (2) system worker.

@@

\paragraph{Scheduling} 
During execution, a worker yields control of its core whenever it performs a
long-latency operation, allowing the processor to remain busy while waiting
for the operation to complete. In addition, a programmer can direct scheduling
explicitly via the \Grappa API calls shown in Figure~\ref{fig:scheduling}. To
minimize yield overhead, the \Grappa scheduler operates entirely in user-space
and does little more than store register state of one task and load that of
another.

@@

Each core in a \Grappa system has its own independent scheduler. Schedulers
have three main responsibilities: servicing system workers that process
communication requests as described in Section~\ref{sec:communication};
rescheduling workers that have been waiting on long-latency operations; and
assigning ready tasks to worker resources that have become idle.

@@

\paragraph{Context switching} 
\Grappa context switches between workers non-preemptively. As with other
cooperative multithreading systems, we treat context switches as function
calls, saving and restoring only the callee-saved state as specified in the
x86-64 ABI\~cite{amd64:abi:2012}. This involves saving six general-purpose
64-bit registers and the stack pointer, as well as the 16-bit x87 floating
point control word and the SSE context/status register. Thus, the minimum
amount of state a cooperative context switch routine must save, according to
the ABI, is 62~bytes.

@@

Since \Grappa keeps a very large number of active workers, their context data
will not fit in cache. Therefore, the scheduler has to perform very careful
prefetching in order to reduce context switch cost to a minimum. We determined
empirically that prefetching the fourth worker in the scheduling order is
sufficient. We also prefetch three cache lines from the stack for each worker.
This makes context switching effectively free of cache misses even to hundreds
of thousands of threads. We provide a detailed analysis of our context switch
performance in Section~\ref{eval:ctx}.

@@

@@

\begin{figure}[htbp]
  \begin{center}
	\begin{tabular}{l}
    \texttt{\scriptsize spawn( void (*fp)(args) )} \\
      Creates a new stealable task \\
    \texttt{\scriptsize spawn\_on( core, (*fp)(args) )} \\
      Creates a new task that will run on a specific core \\
    \texttt{\scriptsize parallel\_for( (*fp)(args), start, end )} \\
      Executes iterations of a loop as stealable tasks  \\
    \texttt{\scriptsize call\_on( core, (*fp)(args) )} \\ 
      Runs a limited function on a specific core without \\
      consuming \Grappa execution resources 
	\end{tabular}
    \begin{minipage}{0.95\columnwidth}
      \caption{\label{fig:expressing-parallelism} \Grappa API: expressing parallelism
      \TODO{update this with latest C++ code from ``api\_examples.cpp''}
      } % \vspace{-4ex}}
    \end{minipage}
    %\vspace{-3ex}
  \end{center}
\end{figure}
